Optimal Verification of Discrete-time Affine Systems

### 2019-04-18 paper.tex
Updates from Overleaf

\paragraph{Related works .} 
The paper solves a formal verification problem. Numerous works (for example~\cite{DBLP:journals/corr/abs-1111-5223,gonnord2014abstract,mine-ESOP16,Sankaranarayanan_BenSassi__2017__Template}) about the resolution of the problem by abstract interpretation are available. They are mainly concerned in the representation of the reachable values. The over-approximations produced are not sufficient to solve a specific verification problem with a given property.

@@

The closest work seems to be the work of Ahmadi et al~\cite{7403149,ahmadi2018robust} which introduces the class of robust-to-dynamics optimization problem. Ahmadi et al solve the problem where the dynamics and the objective function are linear. Moreover, they need a bounded invariant set to solve their problem. In our work, we deal with non-linear quadratic  forms and affine dynamical systems.  Moreover, we do not need a bounded invariant to solve our robust-to-dynamics optimization problem.  Nevertheless, some similarities exist between their approach and the one proposed here : the use of a solution of the discrete Lyapunov matrix equation and the computation of a number of steps to convergence.   
\paragraph{Contributions}
Our very first concern is the resolution of the verification problem for the class of discrete-time affine dynamical systems. We assume that the given property is written as the sublevel set of some quadratic function. 
We reduced the verification problem to an optimization problem :
\begin{equation}
\label{eqintro}
\sup\{f(x)\mid x_{k+1}=g(x_k),\ x_0\in\xin\}=\sup\{f( g^k (x_0))\mid  x_0\in\xin,\ k\in\nn\}\enspace.
\end{equation}
where $f$ is a quadratic function, $g$ an affine function and $\xin$ is a polytope. We remark that the optimal value in~\eqref{eqintro} only depends on the initial values.
%\[
%\sup\{f( g^k (x_0))\mid  x_0\in\xin,\ k\in\nn\}\enspace.
%\]
In the paper, we compute a number of steps $\bigk$ after which the search of the optimal value is useless i.e.
 \begin{equation}
\label{eqintro2}
\sup\{f( g^k (x_0))\mid  x_0\in\xin,\ k\in\nn\}=
\sup\{f( g^k (x_0))\mid  x_0\in\xin,\ k\in \{0,\ldots,\bigk\}\}
\end{equation}
This number of steps $\bigk$ is valid for all possible initial values. We are looking for the best possible integer $\bigk$ for which 
Eq.~\eqref{eqintro2} is true. We propose to construct a family of integers parameterized by some positive reals and some matrices. The approach proposed in the paper uses basic tools from matrix theory such as the discrete Lyapunov matrix equation, minimal and maximal eigenvalues, matrix norms and generalized Raleigh quotients. In theory, the best proposed integer is then the optimal value of some non-convex semi-definite program.  
The exact resolution of this semi-definite problem is, for the moment, left open and we just provide an over-approximation of its optimal value which does not degrade the exact resolution of the verification problem.   For the moment, we propose in practice the computation of two families of integers which over-approximate the best value.

@@

\paragraph{Organization of the paper}
Section~\ref{statement} states the problem i.e. the formulation of our verification problem and its translation into a optimization problem. Section~\ref{mainresults} presents the main results. We give details about the construction of an integer solution of Eq.~\eqref{eqintro2}. Section~\ref{computations} describes the practical idea used in the implementation.  Section~\ref{experiments} is devoted to the experimentations. Finally, Section~\ref{conclusion} concludes and opens future research directions.

@@

\paragraph{Notations}
%In this paper, we denote by $\Idd$ the identity matrix of size $d\times d$. We denote, for a matrix $M$, by $M^\intercal$ the transpose of $M$. The set of the (square) symmetric matrices  (the matrices $M$ such that $M=M^\intercal$) of size $d$ is denoted by $\sym$.
In this paper, we denote by $\Idd$ the identity matrix of size $d$.
We also denote, for a matrix $M$, by $M^\intercal$ the transpose of $M$. The matrix $M$ is {\it symmetric} iff $M=M^\intercal$.

@@

\begin{prop}
\label{propcont}
For all $P\succ 0$, for all $t\in\cont(P)$, $t\geq \lmax{P^{-1/2} Q P^{-1/2}}$.
\end{prop}
%\begin{proof}
%Let $P\succ 0$, let $t\in\cont(P)$. Let $x\in\rd$, $x\neq 0$. We have 
%$x^\intercal Q x\leq t x^\intercal P x$ and thus $(x^\intercal Q x)(x^\intercal P x)^{-1}\leq t$. Finally:
%$\sup_{x\neq 0}(x^\intercal Q x)(x^\intercal P x)^{-1}\leq t $. Writing $y=P^{1/2}x$ ( recall that $P^{1/2}$ is invertible as $P\succ 0$), we obtain :
%$\sup_{y\neq 0}(y^\intercal P^{-1/2}Q P^{-1/2}y)(y^\intercal y)^{-1}\leq t $. However, it is well-know that
%$\sup_{z\neq 0} (z^\intercal M z)(z^\intercal z)^{-1}=\lmax{M}$. In fact, $\lmax{M}$ is achieved for an eigenvector associated with $\lmax{M}$. Hence, 
%$\sup_{y\neq 0} (y^\intercal P^{-1/2}Q P^{-1/2}y)(y^\intercal y)^{-1}=\lmax{P^{-1/2} Q P^{-1/2}}$. Finally $\lmax{P^{-1/2} Q P^{-1/2}}P-Q\succeq 0$ is a direct consequence of Lemma~\ref{lemma1}. 
%
%\end{proof}
To compute our $\bigk$, we introduce a new notation and a new assumption. 
 \begin{assumption}
\label{assumR}
There exists $k\in\nn$ such that\[\min\left\{\sup_{x\in\xin} x^\intercal Q x , \sup_{x\in\xin} x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\right\}\] is strictly positive.
\end{assumption}

@@

\begin{lemma}
\label{lyapnorm}
For all $P\in\lyap{A}$, $0<\norm{A}_P<1$. 
\end{lemma}

%\begin{proof}
%Let $P$ such that $P\succ 0$ and $P-A^\intercal P A\succ 0$.  First, the matrix $A$ is not null then its spectral radius is strictly positive. Then since $\norm{\cdot}_P$ is a matrix norm, $0<\rho(A) \leq \norm{A}_P$.
%
%Secondly, since $P-A^\intercal P A\succ 0$ and using Eq.~\eqref{normdef}, we have $\norm{A}_P\leq 1$. Still using Eq.~\eqref{normdef}, to prove $\norm{A}_P<1$, it suffices to exhibit $\varepsilon>0$ such that $(1-\varepsilon) P-A^\intercal P A\succeq 0$ with $0<\varepsilon<1$. 
%
%Let us write $\varepsilon=\lmin{P-A^\intercal P A} \lmax{P}^{-1}$. Let us prove that $0<\varepsilon<1$ and $(1-\varepsilon) P-A^\intercal P A\succeq 0$.  Since $P-A^\intercal P A\succ 0$, $\lmin{P-A^\intercal P A}>0$ and from $P\succ 0$ we conclude that $\varepsilon >0$. Since $P=P-A^ \intercal P A +A^\intercal P A$, we have, from Lemma~\ref{lemmaWeyl}, $\lmin{P-A^\intercal P A}+\lmax{A^\intercal P A}\leq \lmax{P}$. Now, $A^\intercal P A\succeq 0$ from $P\succ 0$ and then $\lmax{A^\intercal P A}\geq 0$. We conclude that $\varepsilon\leq 1$. Now, if $\lmin{P-A^\intercal P A}=\lmax{P}$ i.e. $\varepsilon=1$, we have 
%$\lmax{A^\intercal P A}=0$ and then $x^\intercal A^\intercal P A x=0$ for all $x\in \rd$ and $A$ is the null matrix. 
%Finally, $\varepsilon<1$. 
%
%Now from Lemma~\ref{lemma1}, $\varepsilon P=P\lmax{P}^{-1}\lmin{P-A^\intercal P A}\preceq \lmin{P-A^\intercal P A}Id_d\preceq 
%P-A^\intercal P A$ which implies that $(1-\varepsilon) P -A^\intercal P A\succeq 0$. 
%\end{proof}

@@

\begin{lemma}
\label{numerator}
For all $t>0$ and $P\in\lyap{A}$ such that $P\in\lyat{A,Q}$: \[
\left(\lfrak+\vqQ(t,P)^2)^{1/2}-\vqQ(t,P)\right) t^{-1/2}\mu(P)^{-1}\in (0,1]\enspace .
\]
\end{lemma}

%\begin{proof}
%First $t>0$ and $P\succ 0$ then $t^{-1/2}\mu(P)^{-1}>0$ and $\vqQ(t,P)\geq 0$. Using Assumption~\ref{assumR}, we get $(\lfrak+\vqQ(t,P)^2)^{1/2}-\vqQ(t,P)> \vqQ(t,P)-\vqQ(t,P)=0$. We conclude that $\left((\lfrak+\vqQ(t,P)^2)^{1/2}-\vqQ(t,P)\right) t^{-1/2}\mu(P)^{-1}>0$. 
%
%Now, since $\lfrak$ and $\vqQ(t,P)$ are non-negative and for all $a,b\geq 0$, $\sqrt{a+b}\leq \sqrt{a}+\sqrt{b}$  then : 
%$(\lfrak+\vqQ(t,P)^2)^{1/2}\leq \lfrak^{1/2}+\vqQ(t,P)$. Hence : 
%\[
%\begin{array}{lll}
%&\left((\lfrak+\vqQ(t,P)^2)^{1/2}-\vqQ(t,P)\right) t^{-1/2}\mu(P)^{-1}& \\
%\leq &\lfrak^{1/2} t^{-1/2}\mu(P)^{-1} & \\
%\leq &(\displaystyle{\sup_{x\in\xin} x^\intercal Q x})^{1/2} t^{-1/2}\mu(P)^{-1} & \text{ from the def. of } \lfrak
%\end{array}
%\]
%
%Since $t>0$ and $P\in\lyap{A}$ satisfy $P\in\lyat{A,Q}$, then $t P-Q\succeq 0$. It follows that 
%$t x^\intercal P x\geq x^\intercal Q x$ for all $x\in \xin$. Taking the supremum over $\xin$ leads to $(\sup_{x\in\xin} x^\intercal Q x)\mu(P)^{-2}t^{-1} \leq 1$. 
%\end{proof}

@@

%\begin{proof}
%Let us prove the first assertion.
%Let $P\in\lyap{A}$. Let $t\in\cont(P)$. The integer part is increasing then we have to prove that the argument of the integer part is increasing in $t$. From Lemma~\ref{lemma2}, we know that $\ln{\norm{A}_P}<0$. Hence, since the natural logarithm is increasing it suffices to show that $\varphi:t:\mapsto \left(\sqr{\lfrak+\vqQ(t,P)^2}-\vqQ(t,P)\right)\left(\sqr{t}\mu(P)\right)^{-1}$ is decreasing. However :
%\[
%\begin{array}{lll}
%\varphi(t)&=&\dfrac{\lfrak}{\left(\sqr{\lfrak+\vqQ(t,P)^2}+\vqQ(t,P)\right)\sqr{t}\mu(P)}\\
%\\
%&=&\dfrac{\lfrak}{\left(\sqr{\lfrak+\dfrac{\norm{q}_2^2 }{4t\lmin{P}}}+\dfrac{\norm{q}_2 }{2\sqr{t\lmin{P}}}\right)\sqr{t}\mu(P)}\\
%\\
%&=&\dfrac{\lfrak}{\left(\sqr{t\lfrak+\dfrac{\norm{q}_2^2 }{4\lmin{P}}}+\dfrac{\norm{q}_2 }{2\sqr{\lmin{P}}}\right)\mu(P)}
%\end{array}
%\]
%We conclude that $t\mapsto \varphi(t)$ is decreasing as an inverse of an increasing function. 
%Finally, Eq.~\eqref{infsolved} follows from Prop.~\ref{propcont} and $t\mapsto K(t,P)$ is increasing for all $P\in\lyap{A}$.  
%
%Now, let $t>0$ and $P\in\lyat{A,Q}$. To prove $K(t,P)=K(1,tP)$ we have to show that $\vqQ(t,P)=\vqQ(1,tP)$, $\sqrt{t}\mu(P)=\mu(tP)$
%and $\norm{A}_{tP}=\norm{A}_P$. The fact $\vqQ(t,P)=\vqQ(1,tP)$ comes from the fact that $\lmin{tP}=t\lmin{P}$ since $t>0$. Second the supremum is homogeneous and then $\sqrt{t}\mu(P)=\mu(tP)$. Finally, $\norm{A}_{tP}=\norm{A}_P$ is a direct consequence of the definition given at Eq.~\eqref{normdef}. We conclude that :
%\[
%\begin{array}{llll}
%\displaystyle{\inf_{t>0} \inf_{P\in\lyat{A,Q}} K(t,P)}&=&\displaystyle{\inf_{t>0} \inf_{P\in\lyat{A,Q}} K(1,tP)}& \text{ from } K(t,P)=K(1,tP)\\
%&=&\displaystyle{\inf_{t>0} \inf_{tP\in\lyaun} K(1,tP)}&\text{ from Eq.~\eqref{simpleequiv}} \\ 
%&=&\displaystyle{\inf_{t>0} \inf_{P\in\lyaun} K(1,P)}&\\
%&=&\displaystyle{\inf_{P\in\lyaun} K(1,P)}&
%\end{array}
%\]
%\end{proof}

@@

We now present the main result of the paper.
\begin{theorem}
\label{thfond}
Let us define \[\bigk=\min\left\{\inf_{P\in\lyaun} K(1,P),\inf_{P\in\lyap{A}} K(\lmax{P^{-1/2}QP^{-1/2}},P)\right\}\enspace .\] Let $k\geq \bigk$. Then : 
\[
\max_{x\in \xin} x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\leq \lfrak
\]
\end{theorem}

%\begin{proof}
%Let $t>0$ and $P\in\lyap{A}$ such that $P\in\lyat{A,Q}$. Let also $k\in\nn$ and $x\in\xin$. We have:
%\[
%x^\intercal (A^k)^\intercal Q A^k x  
%\leq  t \norm{A^k x}_P^2\leq t\norm{A}^{2k}_P \norm{x}^2_P
%\leq t\norm{A}_P^{2k} \mu(P)^2
%\]
%The first inequality comes from the definition of $\lyat{A,Q}$, the second from the matrix norm definition and the last from the definition of $\mu(P)$ (see Eq.~\eqref{intgfond}).
%
%We also have, for all $x\in\xin$ :
%\[
%q^\intercal A^k x
%\leq \norm{q}_2 \norm{A^k x}_2\leq \dfrac{\norm{q}_2 \norm{A^k x}_P}{\sqr{\lmin{P}}} 
%\leq \dfrac{\norm{q}_2\norm{A}_P^{k} \mu(P)}{\sqr{\lmin{P}}}
%\]
%The first inequality comes from Cauchy-Schwarz, the second from Lemma~\ref{lemma1} and the last from the matrix norm definition and the definition of $\mu(P)$. 
%
%Summing the two parts leads to :
%\[
%x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\leq \left(\sqrt{t}\norm{A}_P^k\mu(P)+\vqQ(t,P)\right)^2-\vqQ(t,P)^2
%\]
%where $\vqQ$ is defined at Eq.~\eqref{intgfond}. Then, $ \left(\sqrt{t}\norm{A}_P^k\mu(P)+\vqQ(t,P)\right)^2-\vqQ(t,P)^2\leq \lfrak$ implies that $ \max_{x\in \xin} x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\leq \lfrak$. Now, we remark that  $\left(\sqrt{t}\norm{A}_P^k\mu(P)+\vqQ(t,P)\right)^2-\vqQ(t,P)^2\leq \lfrak$ is equivalent to :
%\[
%\norm{A}_P^k\leq \left(\sqr{\lfrak+\vqQ(t,P)^2}-\vqQ(t,P)\right)\left(\sqr{t}\mu(P)\right)^{-1}
%\]
%Using the natural logarithm and Lemma~\ref{lyapnorm}, the condition  $k\geq K(t,P)$ is sufficient for the latter inequality .  The latter proof is valid for all couple $t>0$ and $P\in\lyap{A}$ such that $P\in\lyat{A,Q}$. So it remains true for   $\btilde>0$ and $\bP\in\lyap{A}$ such that $P\in\lyat{A,Q}$ and $\bigk=K(\btilde,\bP)$. 
%\end{proof}
We recall that $\lfrak$ is smaller than $\max_{x\in\xin} x^\intercal {A^{\ks}}^\intercal Q A^{\ks} x+q^\intercal A^{\ks} x$. Hence, the optimal value of Problem~\ref{pbopt} is greater than $\lfrak$ since $A^{\ks} (\xin)\subseteq \rea$. We conclude that, following Th.~\ref{thfond}, the optimal value of Problem~\ref{pbopt} is attained for powers of $A$ between $\ks$  and $\max\{\bigk-1,\ks\}$.

@@

We recall Weyl's inequalities~\cite{horn1990matrix} that provide inequalities for the eigenvalues of the sum of two symmetric matrices and the sum of the eigenvalues of each matrix.  
\begin{lemma}[Weyl's inequalities]
\label{lemmaWeyl}
Let $M$ and $N$ be symmetric matrices. We have, for all $\ell\in\{1,\ldots,d\}$ :
\[
\lambda_\ell(M)+\lmin{N}\leq \lambda_\ell(M+N)\leq \lambda_\ell(M)+\lmax{N}
\]
\end{lemma}

>>>>

\paragraph{Related works.} 
The paper solves a formal verification problem. Numerous works (for example~\cite{DBLP:journals/corr/abs-1111-5223,gonnord2014abstract,mine-ESOP16,Sankaranarayanan_BenSassi__2017__Template}) about the resolution of the problem by abstract interpretation are available. They are mainly concerned in the representation of the reachable values. The over-approximations produced are not sufficient to solve a specific verification problem with a given property.

@@

The closest work seems to be the work of Ahmadi et al~\cite{7403149,ahmadi2018robust} which introduces the class of robust-to-dynamics optimization problem. Ahmadi et al solve the problem where the dynamics and the objective function are linear. Moreover, they need a bounded invariant set to solve their problem. In our work, we deal with non-linear quadratic  forms and affine dynamical systems.  Moreover, we do not need a bounded invariant to solve our robust-to-dynamics optimization problem.  Nevertheless, some similarities exist between their approach and the one proposed here : the use of a solution of the discrete Lyapunov matrix equation and the computation of a number of steps to convergence.   
\paragraph{Contributions.}
Our very first concern is the resolution of the verification problem for the class of discrete-time affine dynamical systems. We assume that the given property is written as the sublevel set of some quadratic function. 
We reduced the verification problem to an optimization problem :
\begin{equation}
\label{eqintro}
\sup\{f(x)\mid x_{k+1}=g(x_k),\ x_0\in\xin\}=\sup\{f( g^k (x_0))\mid  x_0\in\xin,\ k\in\nn\}\enspace.
\end{equation}
where $f$ is a quadratic function, $g$ an affine function and $\xin$ is a polytope. We remark that the optimal value in~\eqref{eqintro} only depends on the initial values.
%\[
%\sup\{f( g^k (x_0))\mid  x_0\in\xin,\ k\in\nn\}\enspace.
%\]
In the paper, we compute a number of steps $\bigk$ after which the search of the optimal value is useless i.e.
 \begin{equation}
\label{eqintro2}
\sup\{f( g^k (x_0))\mid  x_0\in\xin,\ k\in\nn\}=
\sup\{f( g^k (x_0))\mid  x_0\in\xin,\ k\in \{0,\ldots,\bigk\}\}
\end{equation}
This number of steps $\bigk$ is valid for all possible initial values. We are looking for the best possible integer $\bigk$ for which 
Eq.~\eqref{eqintro2} is true. We propose to construct a family of integers parameterized by some positive reals and some matrices. The approach proposed in the paper uses basic tools from matrix theory such as the discrete Lyapunov matrix equation, minimal and maximal eigenvalues, matrix norms and generalized Raleigh quotients. In theory, the best proposed integer is then the optimal value of some non-convex semi-definite program.  
The exact resolution of this semi-definite problem is, for the moment, left open and we just provide an over-approximation of its optimal value which does not degrade the exact resolution of the verification problem.   For the moment, we propose in practice the computation of two families of integers which over-approximate the best value.

@@

\paragraph{Organization of the paper.}
Section~\ref{statement} states the problem i.e. the formulation of our verification problem and its translation into a optimization problem. Section~\ref{mainresults} presents the main results. We give details about the construction of an integer solution of Eq.~\eqref{eqintro2}. Section~\ref{computations} describes the practical idea used in the implementation.  Section~\ref{experiments} is devoted to the experimentations. Finally, Section~\ref{conclusion} concludes and opens future research directions.

@@

\paragraph{Notations.}
%In this paper, we denote by $\Idd$ the identity matrix of size $d\times d$. We denote, for a matrix $M$, by $M^\intercal$ the transpose of $M$. The set of the (square) symmetric matrices  (the matrices $M$ such that $M=M^\intercal$) of size $d$ is denoted by $\sym$.
In this paper, we denote by $\Idd$ the identity matrix of size $d$.
We also denote, for a matrix $M$, by $M^\intercal$ the transpose of $M$. The matrix $M$ is {\it symmetric} iff $M=M^\intercal$.

@@

\begin{prop}
\label{propcont}
For all $P\succ 0$, for all $t\in\cont(P)$, $t\geq \lmax{P^{-1/2} Q P^{-1/2}}$.
\end{prop}
\begin{proof}
In appendix.
\end{proof}
%\begin{proof}
%Let $P\succ 0$, let $t\in\cont(P)$. Let $x\in\rd$, $x\neq 0$. We have 
%$x^\intercal Q x\leq t x^\intercal P x$ and thus $(x^\intercal Q x)(x^\intercal P x)^{-1}\leq t$. Finally:
%$\sup_{x\neq 0}(x^\intercal Q x)(x^\intercal P x)^{-1}\leq t $. Writing $y=P^{1/2}x$ ( recall that $P^{1/2}$ is invertible as $P\succ 0$), we obtain :
%$\sup_{y\neq 0}(y^\intercal P^{-1/2}Q P^{-1/2}y)(y^\intercal y)^{-1}\leq t $. However, it is well-know that
%$\sup_{z\neq 0} (z^\intercal M z)(z^\intercal z)^{-1}=\lmax{M}$. In fact, $\lmax{M}$ is achieved for an eigenvector associated with $\lmax{M}$. Hence, 
%$\sup_{y\neq 0} (y^\intercal P^{-1/2}Q P^{-1/2}y)(y^\intercal y)^{-1}=\lmax{P^{-1/2} Q P^{-1/2}}$. Finally $\lmax{P^{-1/2} Q P^{-1/2}}P-Q\succeq 0$ is a direct consequence of Lemma~\ref{lemma1}. 
%
%\end{proof}
To compute our $\bigk$, we introduce a new notation and a new assumption. 
 \begin{assumption}
\label{assumR}
There exists $k\in\nn$ such that\[\min\left\{\sup_{x\in\xin} x^\intercal Q x , \sup_{x\in\xin} x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\right\}\] is strictly positive.
\end{assumption}

@@

\begin{lemma}
\label{lyapnorm}
For all $P\in\lyap{A}$, $0<\norm{A}_P<1$. 
\end{lemma}
\begin{proof}
In appendix.
\end{proof}
%\begin{proof}
%Let $P$ such that $P\succ 0$ and $P-A^\intercal P A\succ 0$.  First, the matrix $A$ is not null then its spectral radius is strictly positive. Then since $\norm{\cdot}_P$ is a matrix norm, $0<\rho(A) \leq \norm{A}_P$.
%
%Secondly, since $P-A^\intercal P A\succ 0$ and using Eq.~\eqref{normdef}, we have $\norm{A}_P\leq 1$. Still using Eq.~\eqref{normdef}, to prove $\norm{A}_P<1$, it suffices to exhibit $\varepsilon>0$ such that $(1-\varepsilon) P-A^\intercal P A\succeq 0$ with $0<\varepsilon<1$. 
%
%Let us write $\varepsilon=\lmin{P-A^\intercal P A} \lmax{P}^{-1}$. Let us prove that $0<\varepsilon<1$ and $(1-\varepsilon) P-A^\intercal P A\succeq 0$.  Since $P-A^\intercal P A\succ 0$, $\lmin{P-A^\intercal P A}>0$ and from $P\succ 0$ we conclude that $\varepsilon >0$. Since $P=P-A^ \intercal P A +A^\intercal P A$, we have, from Lemma~\ref{lemmaWeyl}, $\lmin{P-A^\intercal P A}+\lmax{A^\intercal P A}\leq \lmax{P}$. Now, $A^\intercal P A\succeq 0$ from $P\succ 0$ and then $\lmax{A^\intercal P A}\geq 0$. We conclude that $\varepsilon\leq 1$. Now, if $\lmin{P-A^\intercal P A}=\lmax{P}$ i.e. $\varepsilon=1$, we have 
%$\lmax{A^\intercal P A}=0$ and then $x^\intercal A^\intercal P A x=0$ for all $x\in \rd$ and $A$ is the null matrix. 
%Finally, $\varepsilon<1$. 
%
%Now from Lemma~\ref{lemma1}, $\varepsilon P=P\lmax{P}^{-1}\lmin{P-A^\intercal P A}\preceq \lmin{P-A^\intercal P A}Id_d\preceq 
%P-A^\intercal P A$ which implies that $(1-\varepsilon) P -A^\intercal P A\succeq 0$. 
%\end{proof}

@@

\begin{lemma}
\label{numerator}
For all $t>0$ and $P\in\lyap{A}$ such that $P\in\lyat{A,Q}$: \[
\left(\lfrak+\vqQ(t,P)^2)^{1/2}-\vqQ(t,P)\right) t^{-1/2}\mu(P)^{-1}\in (0,1]\enspace .
\]
\end{lemma}
\begin{proof}
In appendix.
\end{proof}
%\begin{proof}
%First $t>0$ and $P\succ 0$ then $t^{-1/2}\mu(P)^{-1}>0$ and $\vqQ(t,P)\geq 0$. Using Assumption~\ref{assumR}, we get $(\lfrak+\vqQ(t,P)^2)^{1/2}-\vqQ(t,P)> \vqQ(t,P)-\vqQ(t,P)=0$. We conclude that $\left((\lfrak+\vqQ(t,P)^2)^{1/2}-\vqQ(t,P)\right) t^{-1/2}\mu(P)^{-1}>0$. 
%
%Now, since $\lfrak$ and $\vqQ(t,P)$ are non-negative and for all $a,b\geq 0$, $\sqrt{a+b}\leq \sqrt{a}+\sqrt{b}$  then : 
%$(\lfrak+\vqQ(t,P)^2)^{1/2}\leq \lfrak^{1/2}+\vqQ(t,P)$. Hence : 
%\[
%\begin{array}{lll}
%&\left((\lfrak+\vqQ(t,P)^2)^{1/2}-\vqQ(t,P)\right) t^{-1/2}\mu(P)^{-1}& \\
%\leq &\lfrak^{1/2} t^{-1/2}\mu(P)^{-1} & \\
%\leq &(\displaystyle{\sup_{x\in\xin} x^\intercal Q x})^{1/2} t^{-1/2}\mu(P)^{-1} & \text{ from the def. of } \lfrak
%\end{array}
%\]
%
%Since $t>0$ and $P\in\lyap{A}$ satisfy $P\in\lyat{A,Q}$, then $t P-Q\succeq 0$. It follows that 
%$t x^\intercal P x\geq x^\intercal Q x$ for all $x\in \xin$. Taking the supremum over $\xin$ leads to $(\sup_{x\in\xin} x^\intercal Q x)\mu(P)^{-2}t^{-1} \leq 1$. 
%\end{proof}

@@

\begin{proof}
In appendix.
\end{proof}
%\begin{proof}
%Let us prove the first assertion.
%Let $P\in\lyap{A}$. Let $t\in\cont(P)$. The integer part is increasing then we have to prove that the argument of the integer part is increasing in $t$. From Lemma~\ref{lemma2}, we know that $\ln{\norm{A}_P}<0$. Hence, since the natural logarithm is increasing it suffices to show that $\varphi:t:\mapsto \left(\sqr{\lfrak+\vqQ(t,P)^2}-\vqQ(t,P)\right)\left(\sqr{t}\mu(P)\right)^{-1}$ is decreasing. However :
%\[
%\begin{array}{lll}
%\varphi(t)&=&\dfrac{\lfrak}{\left(\sqr{\lfrak+\vqQ(t,P)^2}+\vqQ(t,P)\right)\sqr{t}\mu(P)}\\
%\\
%&=&\dfrac{\lfrak}{\left(\sqr{\lfrak+\dfrac{\norm{q}_2^2 }{4t\lmin{P}}}+\dfrac{\norm{q}_2 }{2\sqr{t\lmin{P}}}\right)\sqr{t}\mu(P)}\\
%\\
%&=&\dfrac{\lfrak}{\left(\sqr{t\lfrak+\dfrac{\norm{q}_2^2 }{4\lmin{P}}}+\dfrac{\norm{q}_2 }{2\sqr{\lmin{P}}}\right)\mu(P)}
%\end{array}
%\]
%We conclude that $t\mapsto \varphi(t)$ is decreasing as an inverse of an increasing function. 
%Finally, Eq.~\eqref{infsolved} follows from Prop.~\ref{propcont} and $t\mapsto K(t,P)$ is increasing for all $P\in\lyap{A}$.  
%
%Now, let $t>0$ and $P\in\lyat{A,Q}$. To prove $K(t,P)=K(1,tP)$ we have to show that $\vqQ(t,P)=\vqQ(1,tP)$, $\sqrt{t}\mu(P)=\mu(tP)$
%and $\norm{A}_{tP}=\norm{A}_P$. The fact $\vqQ(t,P)=\vqQ(1,tP)$ comes from the fact that $\lmin{tP}=t\lmin{P}$ since $t>0$. Second the supremum is homogeneous and then $\sqrt{t}\mu(P)=\mu(tP)$. Finally, $\norm{A}_{tP}=\norm{A}_P$ is a direct consequence of the definition given at Eq.~\eqref{normdef}. We conclude that :
%\[
%\begin{array}{llll}
%\displaystyle{\inf_{t>0} \inf_{P\in\lyat{A,Q}} K(t,P)}&=&\displaystyle{\inf_{t>0} \inf_{P\in\lyat{A,Q}} K(1,tP)}& \text{ from } K(t,P)=K(1,tP)\\
%&=&\displaystyle{\inf_{t>0} \inf_{tP\in\lyaun} K(1,tP)}&\text{ from Eq.~\eqref{simpleequiv}} \\ 
%&=&\displaystyle{\inf_{t>0} \inf_{P\in\lyaun} K(1,P)}&\\
%&=&\displaystyle{\inf_{P\in\lyaun} K(1,P)}&
%\end{array}
%\]
%\end{proof}

@@

We now present the main result of the paper.
\begin{theorem}
\label{thfond}
Let us define \[\bigk=\min\left\{\inf_{P\in\lyaun} K(1,P),\inf_{P\in\lyap{A}} K(\lmax{P^{-1/2}QP^{-1/2}},P)\right\}\enspace .\] Let $k\geq \bigk$. Then : 
\[
\max_{x\in \xin} x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\leq \lfrak
\]
\end{theorem}
\begin{proof}
In appendix.
\end{proof}
%\begin{proof}
%Let $t>0$ and $P\in\lyap{A}$ such that $P\in\lyat{A,Q}$. Let also $k\in\nn$ and $x\in\xin$. We have:
%\[
%x^\intercal (A^k)^\intercal Q A^k x  
%\leq  t \norm{A^k x}_P^2\leq t\norm{A}^{2k}_P \norm{x}^2_P
%\leq t\norm{A}_P^{2k} \mu(P)^2
%\]
%The first inequality comes from the definition of $\lyat{A,Q}$, the second from the matrix norm definition and the last from the definition of $\mu(P)$ (see Eq.~\eqref{intgfond}).
%
%We also have, for all $x\in\xin$ :
%\[
%q^\intercal A^k x
%\leq \norm{q}_2 \norm{A^k x}_2\leq \dfrac{\norm{q}_2 \norm{A^k x}_P}{\sqr{\lmin{P}}} 
%\leq \dfrac{\norm{q}_2\norm{A}_P^{k} \mu(P)}{\sqr{\lmin{P}}}
%\]
%The first inequality comes from Cauchy-Schwarz, the second from Lemma~\ref{lemma1} and the last from the matrix norm definition and the definition of $\mu(P)$. 
%
%Summing the two parts leads to :
%\[
%x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\leq \left(\sqrt{t}\norm{A}_P^k\mu(P)+\vqQ(t,P)\right)^2-\vqQ(t,P)^2
%\]
%where $\vqQ$ is defined at Eq.~\eqref{intgfond}. Then, $ \left(\sqrt{t}\norm{A}_P^k\mu(P)+\vqQ(t,P)\right)^2-\vqQ(t,P)^2\leq \lfrak$ implies that $ \max_{x\in \xin} x^\intercal (A^k)^\intercal Q A^k x+q^\intercal A^k x\leq \lfrak$. Now, we remark that  $\left(\sqrt{t}\norm{A}_P^k\mu(P)+\vqQ(t,P)\right)^2-\vqQ(t,P)^2\leq \lfrak$ is equivalent to :
%\[
%\norm{A}_P^k\leq \left(\sqr{\lfrak+\vqQ(t,P)^2}-\vqQ(t,P)\right)\left(\sqr{t}\mu(P)\right)^{-1}
%\]
%Using the natural logarithm and Lemma~\ref{lyapnorm}, the condition  $k\geq K(t,P)$ is sufficient for the latter inequality .  The latter proof is valid for all couple $t>0$ and $P\in\lyap{A}$ such that $P\in\lyat{A,Q}$. So it remains true for   $\btilde>0$ and $\bP\in\lyap{A}$ such that $P\in\lyat{A,Q}$ and $\bigk=K(\btilde,\bP)$. 
%\end{proof}
We recall that $\lfrak$ is smaller than $\max_{x\in\xin} x^\intercal {A^{\ks}}^\intercal Q A^{\ks} x+q^\intercal A^{\ks} x$. Hence, the optimal value of Problem~\ref{pbopt} is greater than $\lfrak$ since $A^{\ks} (\xin)\subseteq \rea$. We conclude that, following Th.~\ref{thfond}, the optimal value of Problem~\ref{pbopt} is attained for powers of $A$ between $\ks$  and $\max\{\bigk-1,\ks\}$.

@@

%We recall Weyl's inequalities
%~\cite{horn1990matrix} 
We recall Weyl's inequalities that provide inequalities for the eigenvalues of the sum of two symmetric matrices and the sum of the eigenvalues of each matrix.  
\begin{lemma}[Weyl's inequalities]
\label{lemmaWeyl}
Let $M$ and $N$ be symmetric matrices. We have, for all $\ell\in\{1,\ldots,d\}$ :
\[
\lambda_\ell(M)+\lmin{N}\leq \lambda_\ell(M+N)\leq \lambda_\ell(M)+\lmax{N}
\]
\end{lemma}

Here is the proof of Proposition~\ref{propcont}.
